CrypTool Script\footnote{ Background reading for the CrypTool program}\\ Mathematics and Cryptography

### 2003-06-30 script/en/chapters/aboutcryptool.tex
richard's corrections

The authors Bernhard Esslinger, Matthias B\"uger, Bartol Filipovic, 
Henrik Koy, Roger Oyono and J\"org Cornelius Schneider
would like to take this opportunity to thank their colleagues 
in the company and at the universities of Frankfurt, Gie\ss en, 
Siegen, Karlsruhe and Darmstadt. They are particularly indepted to Dr.~Peer
Wichmann from the Karlsruhe computer science research centre (Forschungszentrum
Informatik, FZI) for his down-to-earth support.

>>>>

The authors Bernhard Esslinger, Matthias B\"uger, Bartol Filipovic, 
Henrik Koy, Roger Oyono and J\"org Cornelius Schneider
would like to take this opportunity to thank their colleagues 
in the company and at the universities of Frankfurt, Gie\ss en, 
Siegen, Karlsruhe and Darmstadt. They are particularly indebted to Dr.~Peer
Wichmann from the Karlsruhe computer science research centre (Forschungszentrum
Informatik, FZI) for his down-to-earth support.

### 2003-06-30 script/en/chapters/introduction.tex
richard's corrections

In order to also enable developers/authors to work together independently 
the topics have been split up and for each topic an extra chapter has been 
written which can be read on its own. The later editorial work in TeX added 
cross linkages between different sections and footnotes describing where you
can find the according functions within the CrypTool\index{CrypTool} 
program \hyperlink{appendix-menutree}{(see menu tree in appendix A).}
% in \ref{s:appendix-menutree}   AUFPASSEN, DASS nichts doppelt !!
% \hypertarget{appendix-menutree}{}\label{s:appendix-menutree}
Naturally there are much more interesting topics in mathematics and
cryptography which could be discussed in greater depth -- therefore this
is only one of many ways to do it.

\vspace{12pt}
Rapid spread of the Internet has also lead to intensified research in the
technologies involved, especially within the area of cryptography where a good
deal of new knowledge has arisen.

@@

At this point I'd like to thank explicitly 3 people who especially 
contributed to CrypTool\index{CrypTool}. Without their talents and engagement 
CrypTool would not be what it is today:
\vspace{-7pt}
\begin{itemize}
   \item Mr.\ Henrik Koy
   \item Mr.\ J\"org Cornelius Schneider and
   \item Dr.\ Peer Wichmann.
\end{itemize}
Also I want to thank all the many people not mentioned here for their 
hard work (mostly carried out in their spare time).
\\

>>>>

In order to also enable developers/authors to work together independently 
the topics have been split up and for each topic an extra chapter has been 
written which can be read on its own. The later editorial work in TeX added 
cross linkages between different sections and footnotes describing where you
can find the according functions within the CrypTool\index{CrypTool} 
program \hyperlink{appendix-menutree}{(see menu tree in appendix A).}
% in \ref{s:appendix-menutree}   AUFPASSEN, DASS nichts doppelt !!
% \hypertarget{appendix-menutree}{}\label{s:appendix-menutree}
Naturally there are many more interesting topics in mathematics and
cryptography which could be discussed in greater depth -- therefore this
is only one of many ways to do it.

\vspace{12pt}
The rapid spread of the Internet has also lead to intensified research in the
technologies involved, especially within the area of cryptography where a good
deal of new knowledge has arisen.

@@

At this point I'd like to thank explicitly 3 people who particularly 
contributed to CrypTool\index{CrypTool}. Without their talents and engagement 
CrypTool would not be what it is today:
\vspace{-7pt}
\begin{itemize}
   \item Mr.\ Henrik Koy
   \item Mr.\ J\"org Cornelius Schneider and
   \item Dr.\ Peer Wichmann.
\end{itemize}
Also I want to thank all the many people not mentioned here for their 
hard work (mostly carried out in their spare time).
\\

### 2003-06-30 script/en/chapters/numbertheory.tex
richard's corrections

% --------------------------------------------------------------------------
\vskip +40pt
\subsubsection{Further current research in factorisation and number theory}
Prime numbers are part of very many topical research areas in number theory and computer science. 
Progress made with factorisation is bigger than estimated 5 years ago -- this
is not only founded on faster computers but also in new knowledge.

@@

Thus, any progress in efficiency of factorising large integers will effect
the security of the RSA. As a consequence, the underlying primes $p$, $q$
and, thus, the module n (1024 bit as of today) have to be increased. In
the worst case of a breakthrough in factorisation, the RSA algorithm might
be compromised.

@@

Although distributed computing over a large network might help to overcome
this problem, realistic costs for data transfer has to be taken into
account --- a point which was not addressed in Bernstein's proposal.

@@

When this section was written (June 2002) nothing was publicly known,
how far there exist implementations of his theoretical onsets and how
much financing there was for his research project.

@@

Their abstract summarizes their results very well: ``The security of the RSA
cryptosystem depends on the difficulty of factoring large integers. The best
current factoring algorithm is the Number Field Sieve (NFS), and its most
difficult part is the sieving step. In 1999 a large distributed computation
involving thousands of workstations working for many months managed to factor a
512-bit RSA key, but 1024-bit keys were believed to be safe for the next 15-20
years. In this paper we describe a new hardware implementation of the NFS
sieving step ... which is 3-4 orders of magnitude more cost effective than the
best previously published designs ... . Based on a detailed analysis of all the
critical components (but without an actual implementation), we believe that the
NFS sieving step for 1024-bit RSA keys can be completed in less than a year with
a \$10M device, and that the NFS sieving step for 512-bit RSA keys can be
completed in less than ten minutes with a \$10K device. Coupled with recent
results about the difficulty of the NFS matrix step ... this raises some
concerns about the security of these key sizes.''

@@

On January 18, 2002 researchers of the German University of Bonn\footnote{%
\href{http://www.ercim.org/publication/Ercim\_News/enw49/franke.html, 2002-01}
{\texttt{http://www.ercim.org/publication/Ercim\_News/enw49/franke.html, 2002-01}}}
factorised a 158-digit number into its both prime factors (these are build
with 73 and 86 decimal digits) using the GNFS method (General Number Field
Sieve)\index{General Number Field Sieve (GNFS)}.

@@

On January 18, 2002 researchers of the German University of Bonn\footnote{%
   \href{http://www.loria.fr/~zimmerma/records/rsa160}
{\texttt{http://www.loria.fr/\~{}zimmerma/records/rsa160}}} 
factorised a 160-digit number into its both prime factors (these are build
with each 80 decimal digits) using the GNFS method (General Number Field
Sieve)\index{General Number Field Sieve (GNFS)}.

The  computations for the factorization of RSA-160 took place at the 
German Information Security Agency (GISA) in Bonn\footnote{%
Every year the GISA\index{GISA} creates a paper to describe which 
crypto algorithms are feasible to generate digital signatures according
to the German signature law -- under participation of experts from
economy and science. To review signature methods based on the
factorisation problem the GISA also co-operates with researchers of the
University of Bonn.
Further information about crypto algorithms on the web page of GISA:
   \href{http://www.bsi.bund.de/esig/basics/techbas/krypto/index.htm}
{\texttt{http://www.bsi.bund.de/esig/basics/techbas/krypto/index.htm}}.
}.

>>>>

% --------------------------------------------------------------------------
\vskip +40pt
\subsubsection{Further current research in factorisation and number theory}
Prime numbers are part of very many topical research areas in number theory and computer science. 
Progress made with factorisation is bigger than was estimated 5 years ago -- this
is not only due to faster computers but also new knowledge.

@@

Thus, any progress in efficiency of factorising large integers will effect the
security of the RSA. As a consequence, the underlying primes $p$, $q$ and,
thus, the module n (1024 bit as of today) have to be increased. In case of a
quantum leap in factorisation, the RSA algorithm might be compromised.

@@

Although distributed computing over a large network might help to overcome
this problem, realistic costs for data transfer have to be taken into
account --- a point which was not addressed in Bernstein's proposal.

@@

When this section was written (June 2002) nothing was publicly known about, how
far there exist implementations of his theoretical onsets and how much
financing there was for his research project.

@@

Their abstract summarizes their results very well: ``The security of the RSA
cryptosystem depends on the difficulty in factoring large integers. The best
current factoring algorithm is the Number Field Sieve (NFS), and its most
difficult part is the sieving step. In 1999 a large distributed computation
involving thousands of workstations working for many months managed to factor a
512-bit RSA key, but 1024-bit keys were believed to be safe for the next 15-20
years. In this paper we describe a new hardware implementation of the NFS
sieving step ... which is 3-4 orders of magnitude more cost effective than the
best previously published designs ... . Based on a detailed analysis of all the
critical components (but without an actual implementation), we believe that the
NFS sieving step for 1024-bit RSA keys can be completed in less than a year with
a \$10M device, and that the NFS sieving step for 512-bit RSA keys can be
completed in less than ten minutes with a \$10K device. Coupled with recent
results about the difficulty of the NFS matrix step ... this raises some
concerns about the security of these key sizes.''

@@

On January 18, 2002 researchers at the German University of Bonn\footnote{%
\href{http://www.ercim.org/publication/Ercim\_News/enw49/franke.html, 2002-01}
{\texttt{http://www.ercim.org/publication/Ercim\_News/enw49/franke.html, 2002-01}}}
factorised a 158-digit number into its both prime factors (these are build
with 73 and 86 decimal digits) using the GNFS method (General Number Field
Sieve)\index{General Number Field Sieve (GNFS)}.

@@

On January 18, 2002 researchers at the German University of Bonn\footnote{%
   \href{http://www.loria.fr/~zimmerma/records/rsa160}
{\texttt{http://www.loria.fr/\~{}zimmerma/records/rsa160}}} 
factorised a 160-digit number into its both prime factors (these are build
with each 80 decimal digits) using the GNFS method (General Number Field
Sieve)\index{General Number Field Sieve (GNFS)}.

The  computations for the factorization of RSA-160 took place at the 
German Information Security Agency (GISA) in Bonn\footnote{%
Every year the GISA\index{GISA} creates a paper to describe which 
crypto algorithms are feasible to generate digital signatures according
to the German signature law -- under participation of experts from
economy and science. To review signature methods based on the
factorisation problem the GISA also co-operates with researchers from the
University of Bonn.
Further information about crypto algorithms on the web page of GISA:
   \href{http://www.bsi.bund.de/esig/basics/techbas/krypto/index.htm}
{\texttt{http://www.bsi.bund.de/esig/basics/techbas/krypto/index.htm}}.
}.

### 2003-06-30 script/en/chapters/primes.tex
richard's corrections

% --------------------------------------------------------------------------
\subsubsection[Generalized Fermat numbers $f(b,n) = b^{2^n} + 1$]
              {Generalized Fermat numbers\footnotemark~$f(b,n) = b^{2^n} + 1$}
    \footnotetext{%
      The base of this power is no longer restricted to 2 . \\
      Even more generic would be:  $f(b,c,n) = b^{c^n} \pm 1$
    }
\hypertarget{generalizedFermatprimes}{}
\index{Fermat!number!generalized} 
    Generalized Fermat numbers are more numerous than Mersenne numbers at
    equal size and many of them are waiting to be discovered to fill the
    big gaps between the Mersenne primes already found or still undiscovered.
    Progress in number theory made it possible that numbers, where the
    representation is not limited to the base 2, can be tested at almost
    the same speed than a Mersenne number.

@@

\paragraph{Unproven statements / conjectures about primes}\mbox{}
\begin{itemize}
  \item Christian Goldbach\footnote{%
      Christian Goldbach\index{Goldbach Christian}, German mathematician,
      Mar 18, 1690$-$Nov 20, 1764.
      }
      conjectured: Every even natural number greater than 2 can
      be represented as the sum of two prime numbers.
      Computers have verified\footnote{%
It is generally accepted today, that the Goldbach Conjecture is true, 
i.~e.\ valid for all even natural numbers bigger than $2$.
In 1999, mathematician J\"org Richstein from the computer sciences 
institute at the University of Giessen, studied even numbers up to 
400 billion and found no contradictory example (see
\href{http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}
{\tt http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}).
\index{Richstein 1999}
Nevertheless, this does not provide us with general proof.\\
The fact is that despite all efforts, Goldbach's conjecture has to date
not been proven. This leads one to believe that since the pioneer work of
the Austrian mathematician Kurt G\"odel \index{G\""odel Kurt} is 
well-known, not every true mathematical theorem is provable (see
\href{http://www.mathematik.ch/mathematiker/goedel.html}
{http://www.mathematik.ch/mathematiker/goedel.html}).
Perhaps Goldbach's conjecture was correct, but in any case the proof
will never be found. Conversely, that will presumably also remain unproven.}
      the Goldbach conjecture for all even numbers up to
      $4*10^{14}$ but no general proof has yet been found\footnote{%
The English publisher {\em Faber} and the American publisher {\em Bloomsbury}
issued in 2000 the 1992 published book ``Uncle Petros and Goldbach's 
Conjecture'' by Apostolos Doxiadis.
It's the story of an old maths professor who fails to prove a more than
250 year old puzzle. To boost the sales figures the English and American
publishers have offered a prize of 1 million USD, if someone can prove
the conjecture -- which should be published by 2004 in a well-known
mathematical journal.\\
Surprisingly only British and American citizens are allowed to participate.\\
The theorem which has come closest so far to Goldbach's conjecture
was proved by Chen Jing-Run in 1966 in a way which is somewhat hard
to understand:
Each even integer greater than $2$ is the sum of one prime and of 
the product of two primes. E.g.: $20=5+3*5.$\\
Most of the research about the Goldbach conjecture is collected in 
the book: ``Goldbach Conjecture'', ed. Wang Yuan, 1984, World scientific
Series in Pure Maths, Vol. 4.\\
Especially this conjecture makes it clear, that even today we do not
have a complete understanding of the deeper connections between addition and
multiplication of natural numbers.
}.

>>>>

% --------------------------------------------------------------------------
\subsubsection[Generalized Fermat numbers $f(b,n) = b^{2^n} + 1$]
              {Generalized Fermat numbers\footnotemark~$f(b,n) = b^{2^n} + 1$}
    \footnotetext{%
      The base of this power is no longer restricted to 2 . \\
      Even more generic would be:  $f(b,c,n) = b^{c^n} \pm 1$
    }
\hypertarget{generalizedFermatprimes}{}
\index{Fermat!number!generalized} 
    Generalized Fermat numbers are more numerous than Mersenne numbers of a
    equal size and many of them are waiting to be discovered to fill the
    big gaps between the Mersenne primes already found or still undiscovered.
    Progress in number theory made it possible that numbers, where the
    representation is not limited to the base 2, can be tested at almost
    the same speed than a Mersenne number.

@@

\paragraph{Unproven statements / conjectures about primes}\mbox{}
\begin{itemize}
  \item Christian Goldbach\footnote{%
      Christian Goldbach\index{Goldbach Christian}, German mathematician,
      Mar 18, 1690$-$Nov 20, 1764.
      }
      conjectured: Every even natural number greater than 2 can
      be represented as the sum of two prime numbers.
      Computers have verified\footnote{%
It is generally accepted today, that the Goldbach Conjecture is true, 
i.~e.\ valid for all even natural numbers greater than $2$.
In 1999, mathematician J\"org Richstein from the computer sciences 
institute at the University of Giessen, studied even numbers up to 
400 billion and found no contradictory example (see
\href{http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}
{\tt http://www.informatik.uni-giessen.de/staff/richstein/de/Goldbach.html}).
\index{Richstein 1999}
Nevertheless, this does not provide us with general proof.\\
The fact is that despite all efforts, Goldbach's conjecture has to date
not been proven. This leads one to believe that since the pioneer work of
the Austrian mathematician Kurt G\"odel \index{G\""odel Kurt} is 
well-known, not every true mathematical theorem is provable (see
\href{http://www.mathematik.ch/mathematiker/goedel.html}
{http://www.mathematik.ch/mathematiker/goedel.html}).
Perhaps Goldbach's conjecture was correct, but in any case the proof
will never be found. Conversely, that will presumably also remain unproven.}
      the Goldbach conjecture for all even numbers up to
      $4*10^{14}$ but no general proof has yet been found\footnote{%
The English publisher {\em Faber} and the American publisher {\em Bloomsbury}
issued in 2000 the 1992 published book ``Uncle Petros and Goldbach's 
Conjecture'' by Apostolos Doxiadis.
It's the story of an old maths professor who fails to prove a more than
250 year old puzzle. To boost the sales figures the English and American
publishers have offered a prize of 1 million USD, if someone can prove
the conjecture -- which should be published by 2004 in a well-known
mathematical journal.\\
Surprisingly only British and American citizens are allowed to participate.\\
The theorem which has come closest so far to Goldbach's conjecture
was proved by Chen Jing-Run in 1966 in a way which is somewhat hard
to understand:
Each even integer greater than $2$ is the sum of one prime and of 
the product of two primes. E.g.: $20=5+3*5.$\\
Most of the research about the Goldbach conjecture is collected in 
the book: ``Goldbach Conjecture'', ed. Wang Yuan, 1984, World scientific
Series in Pure Maths, Vol. 4.\\
Especially this conjecture makes it clear, that even today we do not
have a complete understanding of the deeper connections between addition and
multiplication of natural numbers.
}.
